---
title: "Investigating Neolithic pottery use in the Yanyuan Basin in Southwest China using organic residue analysis"
author:
  - Myfirstname Mylastname:
      email: fl@oneplace.edu
      institute: [UofO]
      correspondence: true
  - Yourfirstname Yourlastname:
      email: fl@another.edu
      institute: [UofA]
      correspondence: false
institute:
  - UofO: University of One Place
  - UofA: University of Another Place
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The following code chunk defines some general settings how code chunks should behave. -->

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300)
```

# Introduction
<!-- intro to the topic and research aim-->
Movements of people between geographically distinct areas usually accompany the exchange or introduction of goods, species, and ideas that could result in far-reaching impacts on local societies economically and culturally. In Southwest China, the Hengduan Mountain ranges as natural passes play an important role in the migration of farming communities and the spread of crops into this region and mainland Southeast Asia. Recent archaeological evidence from the Yanyuan Basin, located in the middle Hengduan Mountains, reveals the introduction of cultivated millet and rice simultaneously during the early Neolithic period, similar to the mixed farming found in Neolithic central China [@Chen2015; @Hao2022]. However, how and to what extent the two crops were incorporated into the local cultures is still not fully understood. To explore whether the new species became main food sources during the Neolithic transition in the Yanyuan Basin, we investigated food residues from archaeological vessels at Guijiabao, the oldest Neolithic site in the Yanyuan region so far excavated.  

<!-- sites and research questions-->
As an early Neolithic site, Guijiabao (ca. 4800-4300 BP) is crucial to understanding cultural interactions and the spread of crops into the Yanyuan Basin due to its location at the intersection of the middle Hengduan Mountain ranges, the Sichuan Plateau, and the Yunnan-Guizhou Plateau. The recent evidence at Guijiabao reveals millet as the main crop with little rice according to the results of flotation [@citation], different from the abundant rice grains found at other Neolithic sites in that area. Although the plant remains suggest the possible domestication of millet in the early Neolithic, we still lack direct evidence of culinary practices during the Neolithic transition. To explore the extent of acceptance of new crops and species in the Yanyuan Basin, we ask the questions: What was the culinary practices during the Neolithic transition? Could we identify the evidence of new crops or species incorporated into the local dietary systems in the Early Neolithic period? We answer those questions by examining potsherds collected from Guijiabao using organic residue analysis, which could offer a better understanding of food sources during the Neolithic transition in the Yanyuan Basin. 

This study aims to detect and identify organic residues, particularly lipid residues, that could have been preserved in the ceramic fabric to explore the use of pottery vessels at Guijiabao. Lipid analysis have proven useful in archaeology since lipid molecules can be well preserved over thousand years in the porous fabric of pottery with the protection of clay matrices from microbiological degradation and water leaching [@Evershed1993; @Evershed2008; @Roffet2017]. Lipid residues are the most direct evidence to investigate culinary practices and dietary patterns since specific lipid compounds can be associated with different types of foodstuffs cooked or stored in pottery vessels [@Copley2005]. It is especially informative for human subsistence in the situation where archaeological sites lack organic remains due to acidic sediments [@Kwak2020]. The growing application of organic residue analysis demonstrate it offers a better understanding of culinary practices during the transitional period to agricultural societies [@Craig2011], dietary patterns in

# Geographical and Cultural Background of Guijiabao in Yanyuan Basin
<!--Info about Yanyuan area-->
The Yanyuan area in Sichuan province is a pull-apart basin located in the middle eastern part of the Hengduan Mountain, which is a group of parallel mountain ranges in a north-south direction from the Tibetan Plateau to Yunnan-Guizhou Plateau. Although the ranges isolate areas with their own unique geographic conditions and cultures, river valleys provide pathways for the movements of people and the exchange of goods between cultural groups since the prehistoric period [@Lin2022; @Hao2022]. The archaeological sites in the Yanyuan Basin, therefore, are important to understanding the cultural interactions between southwest China and other regions, including northwest China, the Sichuan area, and northwest Yunnan. 

<!--Info about Guijiabao-->
Among those archaeological sites in Yanyuan, Guijiabao located in the southwestern part of the Basin is the only site with continuous occupation from the early Neolithic period to the historical period. The chronology of Guijiabao can be divided into three temporal phases based on 19 radiocarbon dates from plant remains and a human bone, including the Neolithic period (5000-3700 BP.), the Bronze Age (3200-2700 BP.), and the late North Song to early Ming Dynasties (1200-500 BP.) [@Hao2022]. The Neolithic period can be further separated into two subphases, 5000-4500 cal. BP and 4500-3700 cal. BP, according to the cultural remains and pottery characteristics. Recent archaeological evidence indicates the spread of millet-rice mixed economy to southwest Sichuan can be dated back to around 5,000 BP.
 
# Materials and methods

## Samples
We selected five ceramic fragments (Figure XX) from Guijiabao for organic residue analysis, in which four samples were recovered from the trench 13 excavated in the 2017 field season and one sample was from the 2016 season. To further investigate the pottery use in an earlier period, we focus on pottery vessels from the stratigraphic layer which dates to the early Neolithic period (5000-4500 BP.). During this period, the common vessel forms include jars, urns, bowls, and basins. For the pottery types, cord-marked grey pottery vessels with fine inclusions and cord-marked sand-tempered vessels in reddish-brown color with coarse inclusions are the two dominant types [@Hao2022]. Our samples are mostly fragments from jars of those two major types that were supposed to be used for food serving or storage according to the archaeological context of living area []. 

## Methods
In this study, we detected and identified lipid compounds by gas chromatography mass spectrometry (GC-MS), and quantify lipids by gas chromatography with flame-ionization detection (GC-FID). To distinguish the source of the common fatty acids, palmitic (C16:0) and stearic (C18:0), found in most animals and plants, we determined their stable carbon isotope characteristics by gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS). The stable carbon isotope ratio for individual lipid compounds is powerful indicator of the origins of organic residues since different food categories, such as terrestrial or marine animals and plants, have specific ranges of δ^13^C values [@Evershed1994]. 

We cleaned the surface of each potsherd to prevent external contamination from environmental factors and handling during excavation or post-excavation analysis. Approximately 3.5 grams powdered samples were extracted with a mixture of dichloromethane/methanol solution (9:1 v/v) using accelerated solvent extraction (ASE Dionex 350; at 100 degrees C). We conducted saponification by adding 10 mL 1M KOH in methanol and 1 mL water into the extracts and heated them in an oven at 70 degrees C overnight. Saponified extracts were acidified with 6M hydrochloric acid (HCl), and the acid fraction was extracted using hexane ($3\times10$ mL). A 5% aliquot of each sample was derivatized using 20 uL N,O-bis(trimethylsilyl)trifluoroacetamide (BSTFA) with 1% v/v trimethylchlorosilane (TMCS), and heated at 70 degrees C for one hour. After drying under a stream of nitrogen, the derivitized sample was dissolved in toluene (20 uL) prior to GC-MS. 

The remaining 95% aliquot was fractioned by solid phase extraction (Aminopropyl, NH2, column with 0.5 g sorbent) with 8 mL dichloromethane:Isopropyl alcohol (3:1), 8 mL 4% acetic acid in ethyl ether, and 6 mL methanol respectively. The second fraction was methylated to produce fatty acid methyl esters (FAMEs) using 1 mL dry hexane and 2 mL 10% (v/v) acetyl chloride in dry methanol, and heated at 70 degree C overnight. The FAME derivatives were extracted with hexane ($3\times10$ mL) and blown down with nitrogen stream. About 5% aliquot dissolved in toluene was analyzed by GC-FID and the remaining sample was analyzed by GC-C-IRMS. 

# Results
Lipid residues from our pottery samples show that the saturated medium-chain fatty acids, palmitic (C16:0) and stearic(C18:0), are the two dominant compounds. Those two compounds were normally degraded from animal fat and plant wax that indicates the organic compounds in our samples were highly degraded 


. However, for the sample five, we detected rich compounds with high intensity compared to the other four samples. This might indicate a different source.

To further distinguish the source of fatty acids, we conducted stable carbon isotope analysis for those samples. Here you can see the stable carbon isotope measurements of C16:0 and C18:0 fatty acids. The results suggest a dominance of C3 plant for most pots, and a few shows a mixture of animal fats and C3 plants. The sample 5 where more compounds were preserved shows a source of C4 plant or marine animals instead. We think this is very interesting finding since there was just a few rice grains found at this site compared to the evidence of millets. 

# Discussion
The results show a mixture of C3 plants, such as rice, and terrestrial animals for most pots with one pot showing more C4 plant source or marine source. Our striking finding might suggest that consumption of rice could be part of the daily life at Guijiabao and the adoption of rice agriculture could be as important as millet cultivation which was thought the main crop previously. Our next steps is to increase sample size to include pots from different time periods and different types of pottery to identify any variation through time and also try different methods, such as phytolith analysis to further support our statement about rice cultivation. 

# Conclusion
Our results show a mixture of C3 plants and terrestrial animals for most pots with one pot showing more C4 plant sources. This case study not only indicates the possibility of rice agriculture that has not been identified previously at the site but also demonstrates the first use of lipid analysis for pottery to explore the diet of the late Neolithic period in this region. Our case study not only indicates the possibility of rice agriculture that has not been identified previously at the site but also demonstrates the first use of lipid analysis for pottery to explore the diet of the late Neolithic period in this region.


# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References 

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

<div id="refs"></div>

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies: 

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```
